Multi-scale trend analysis of water quality using error propagation of generalized additive models

### 2021-01-21 manu_draft.Rmd
more methods edits from JH

GAMs smooth the raw data across time to separate variability in the response variable associated with an explanatory factor from noise inherent in the raw data. The simplest GAM for this purpose can be expressed as:

@@

Here `y` is the time-series of interest, such as chlorophyll, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knots used to create the spline.  Log$_{10}$-transformed chlorophyll was used for all analyses to meet assumptions of normally-distributed residuals.

For smoothing in `mgcv`, appropriate smoothness is achieved not by choice of knots but by a penalty on the net curvature of the spline [@Wood04].  Smoothing was determined using generalized cross-validation (GCV, as implemented in `mgcv`), which approximately minimizes out-of-sample prediction error.  To allow GCV (or other alternatives) to work as intended, the number of knots must be sufficiently large.  Results should not be sensitive to the number of knots; if they are, the number of knots should be increased.  Therefore, we choose a sufficiently large number of knots for `num_knots_Y` as 12 times the number of years in the time series that was modeled.  This created the potential to have one knot per month as an approach to both prevent under-fitting the observed data and to accurately estimate the seasonal signal within a year.  In some cases, this upper limit was too large given the observed data and the multiplier was reduced by one until sufficient degrees of freedom were available (i.e., 12 * years, 11 * years, etc.).

The next three spline formulations differ in how spline terms compose a model to smooth the raw data.  For convenience, a linear trend in `cont_year` can be specified explicitly as a separate component.  This would be expressed as:

@@

This model is effectively and mathematically equivalent to model S (Table \@ref(tab:modstrtab)).  The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.  When `cont_year` is included explicitly as a linear term in model SY, the `mgcv` implementation automatically adjusts the basis functions for the spline to exclude the linear term, to avoid over-parameterizing the model.  With model S, the estimated trend in `cont_year` and its uncertainty can be extracted from the fitted spline.  With model SY, it is available more conveniently.  Further, package `mgcv` includes an option `select = TRUE` to penalize linear trends in splines to provide a method for variable selection, such as when numerous splines are included in the model formulation for variables that may or may not be important.  For our approach, this option is not used and all models specify `select = FALSE`.  Details in the supplement explain this justification.

Next, an average within-year cyclic pattern as a separate spline can be separated in the model formulation.  This can be expressed as:

@@

Here `doy` is "day-of-year" (i.e., Julian date, a count starting January 1 for each year), `bs = 'cc'` indicates the spline will be cyclic (constrained to start and end at the same value), and `num_knots_D` is the upper limit for the number of knots for the `doy` spline.  While model SYD is not mathematically equivalent to models S and SY, it should be functionally similar.  In this model, the `doy` spline gives the average pattern of fluctuations within each year.  This changes the interpretation of the `cont_year` spline to represent smoothed deviations from the average within-year pattern.

Models S, SY, and SYD can all potentially extract a similar signal from the raw data (Table \@ref(tab:modstrtab)).  What differs between the models is the allocation of penalties for curvature used to determine smoothness for each spline.  In model SYD, there are separate penalties for the two splines, as compared to S and SY that include penalties only for the `cont_year` spline.  Caution is needed in interpreting differences in fit between model SYD and models S or SY because the penalties for smoothing splines based on curvature are heuristic [@Wood17]. For example, if a lower AIC is achieved in one model compared to another, assuming both use sufficient knots, this may just reflect the outcome of alternative penalization heuristics implied by the different formulations. In the examples here, model SYD achieves nearly identical fits to model S or SY, where the latter by definition also achieve identical fits.

Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will "borrow information" across years.  The model will impute an average seasonal pattern where data are limited. However, a conceptual representation of what an average within-year pattern represents is challenging.  For example, suppose a spring chlorophyll peak is a notable feature in every year.  If the timing is the same in every year whereas the magnitudes vary, then the average within-year pattern will have an average magnitude thta appears as a "typical" year.  However, if the magnitude is the same and the timing varies across years, then the average within-year pattern will not appear typical.  The imputed values will represent a day-by-day average, appearing more spread out in time than the pattern in any single year.

More importantly, the average within-year pattern (and its uncertainty) can be obtained from estimated models S and SY, as well as from model SYD, but using a different approach.  In models S and SY, the average within-year pattern is simply the average across years of the smoothed within-year patterns.  From this perspective, model SYD, like model SY, is a convenience, estimating the average within-year pattern directly as a separate estimated spline.  The average pattern extracted from model S or SY may not be identical to the `doy` spline estimated in model SYD, again due to different smoothing formulations, but in practice they should be similar.

@@

Here, `ti()` specifies a tensor-product spline for a surface that varies smoothly as a function of both `cont_year` and `doy`.  The number of knots is the product of `num_knots_Y_ti` in the `cont_year` axis and `num_knots_D_ti` in the `doy` axis.  In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knots_D_ti` or a sufficiently large value for knots in `s(cont_year)`, but not both to avoid over-parameterization.

Following similar rationale above, the relationship of model SYDI to model S is similar to that of model SYD to model S.  Model SYDI is more mathematically different, but all of the splines use the same inputs to smooth the same data.  The splines in `cont_year` and `doy` will likely not capture as much variation as above because of fewer knots.  The `ti` term represents an interaction by allowing the pattern in `cont_year` to vary by `doy` and vice-versa.  The interaction term in model SYDI provides an appearance that this model is fundamentally different from those provided by the other models.  However, models S, SY, and SYD all allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time-series.  Although model SYDI is the only model that includes an explicit interaction term, all of the models support the interaction conceptually. By providing this term with sufficient knots, the raw data can be fully smoothed with model SYDI to a similar degree as for the other models.  However, this outcome is difficult to achieve for reasons explained below.

>>>>

GAMs smooth the raw data across time to separate variability in the response variable associated with an explanatory factor, indepdent of noise inherent in the raw data. The simplest GAM for this purpose is expressed as:

@@

where `y` is the time-series of interest, such as chlorophyll, `cont_year` is "continuous year", a continuous numerical date (e.g., July 1st 2019 would be 2019.5), `y ~ s(...)` indicates that `y` will be explained by a smoothing spline (in this case of `cont_year`), and `num_knots_Y` is the number of knot or "connections" along the spline that influence curvature.  Log$_{10}$-transformed chlorophyll was used for all analyses to meet assumptions of normally-distributed residuals.

The optimal level of smoothing in `mgcv` is determined by a penalty on the net curvature of the spline [@Wood04].  Smoothing was determined using generalized cross-validation (GCV, as implemented in `mgcv`), which approximately minimizes out-of-sample prediction error.  To allow GCV (or other alternatives) to work as intended, a potential upper limit on the number of knots determined by the analyst must be sufficiently large.  Results should not be sensitive to the number of knots; if they are, the number of knots should be increased.  We chose a sufficiently large number of knots for `num_knots_Y` (Model S) as 12 times the number of years in the time series that was modeled.  This created the potential, as determined through GCV, to have one knot per month as an approach to both prevent under-fitting the observed data and to accurately estimate the seasonal signal within a year.  If the data were too sparse, the model will have insufficient degrees of freedom to fit 12 knots per year, in which case the number of knots was reduced by one knot per year until sufficient degrees of freedom were available (i.e., 12 * years, 11 * years, etc.).

The next three spline formulations (Model SY, SYD, and SYDI) provide progressively increasing complexity in how spline terms compose a model to smooth the raw data.  Model SY describes the time series using a linear trend plus a spline for `cont_year`:

@@

This model is mathematically equivalent to model S (Table \@ref(tab:modstrtab)).  The spline for `cont_year` includes an unpenalized linear trend, so a trend will be estimated in model S.  When `cont_year` is included explicitly as a linear term in model SY, `mgcv` adjusts the basis functions for the spline to exclude the linear term, thereby not over-parameterizing the model. Whereas an estimated linear trend in `cont_year` and its uncertainty can be extracted from the fitted spline in model S, model SY provides this trend directly, given the equivalent result.  Further, package `mgcv` can penalize linear trends in splines to provide a method for variable selection (option `select = TRUE`), such as when numerous splines are included in the model formulation for variables that may or may not be important.  For our approach, this option is not used and all models specify `select = FALSE`.  Details in the supplement explain this justification.

MOdel SYD adds an average within-year cyclic pattern as a separate spline:

@@

where `doy` is "day-of-year" (i.e., Julian date, a count starting January 1 for each year), `bs = 'cc'` indicates that the spline will be cyclic (constrained to start and end at the same value), and `num_knots_D` is the upper limit for the number of knots for the `doy` spline.  While model SYD is not mathematically equivalent to models S and SY, it should produce nearly identical results.  The `doy` spline in model SYD gives the average seasonal pattern and changes the interpretation of the `cont_year` spline to represent smoothed deviations from the average within-year pattern.

Models S, SY, and SYD can all potentially extract a similar signal from the raw data (Table \@ref(tab:modstrtab)).  What differs between the models is the allocation of penalties for curvature used to determine smoothness for each spline.  In model SYD, there are separate penalties for the two splines, as compared to S and SY that include penalties only for the `cont_year` spline. This is important because variation in the response variable can be differently attributed to each spline depending on model, whereas the sum of components for each model produce comparable results between models.  Our goal is to extract signals (i.e., seasonal estimates) from the fitted time series and different allocation of penalties among the splines in each model is unimportant for this task.  However, interpreting differences in fit between model SYD and models S or SY can be problematic because the penalties for smoothing splines based on curvature are heuristic [@Wood17]. For example, if a lower AIC is achieved in one model compared to another, assuming both use sufficient knots, this may just reflect the outcome of alternative penalization heuristics implied by the different formulations and does not imply one model fits better. In the examples here, model SYD achieves nearly identical fits to model S or SY, where the latter by definition also achieve identical fits.

Model SYD has the appealing feature that, if some parts of some years have limited data, model SYD will consider data from the same period in other years by imputing an average seasonal pattern with the `doy` spline. However, an interpretation of these imputations is challenging.  For example, the spring chlorophyll peak is a notable feature every year in south SFE.  If the peak occurs at the same time every year but the magnitude varies, then the average within-year pattern can be interpreted as the average magnitude. However, if the magnitude is the same but the timing varies across years, then the magnitude of the average peak cannot be similarly interpreted and instead underestimates the magnitude that usually occurs.  Moreover, the width or duration of the peak will be longer than typically occurs in a given year.

@@

where `ti()` specifies a tensor-product spline for a surface that varies smoothly as a function of both `cont_year` and `doy`.  The number of knots is the product of `num_knots_Y_ti` in the `cont_year` axis and `num_knots_D_ti` in the `doy` axis.  In SYDI, the need for sufficient knots can be satisfied either by sufficiently large values for `num_knots_Y_ti` and `num_knots_D_ti` or a sufficiently large value for knots in `s(cont_year)`, but not both given limits on the model degrees of freedom.

Following the rationale above, the relationship of model SYDI to model S is similar to that of model SYD to model S.  Model SYDI differs formulaically from model S to a greater extent than model SYDI, but all of the splines use the same inputs to smooth the same data.  The splines in `cont_year` and `doy` will likely not capture as much variation in model SYDI compared to model SY given the fewer knots that are available to the former.  The `ti` term represents an interaction by allowing the pattern in `cont_year` to vary by `doy` and vice-versa.  The interaction term in model SYDI provides an appearance that this model is fundamentally different from those provided by the other models.  However, models S, SY, and SYD all allow within-year fluctuations to vary across years by allowing a spline to be fit through the entire time-series.  Although model SYDI is the only model that includes an explicit interaction term, all of the models support the interaction conceptually. By providing this term with sufficient knots, the raw data can be fully smoothed with model SYDI to a similar degree as for the other models.  However, this outcome is difficult to achieve for reasons explained below.
